---
title: "Renal Disease: NAPAP and Serum Creatinine Analysis"
subtitle: "Problems 12.47 - 12.51"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 8)
```

# Overview

This analysis examines the relationship between NAPAP (N-Acetyl-p-aminophenol, also known as acetaminophen or paracetamol) exposure and serum creatinine levels as a marker of renal function. The data come from a longitudinal study of Swiss women measured from 1968 to 1978.

## Study Groups

- **Group 1**: High-NAPAP exposure
- **Group 2**: Low-NAPAP exposure
- **Group 3**: Control group (no NAPAP exposure)

## Research Questions

**Problem 12.47**: Compare mean change in serum creatinine from baseline (1968) to 1978 among the three groups using ANOVA.

**Problem 12.48**: Fit individual regression lines (creatinine vs. time) for each person in each group and interpret slopes and intercepts.

**Problem 12.49**: Compare the slopes of regression lines among the three groups.

**Problem 12.50**: Compare the intercepts of regression lines among the three groups.

**Problem 12.51**: Provide overall conclusions about serum creatinine differences among groups.

# Load Required Libraries

```{r load-libraries}
library(tidyverse)
library(knitr)
library(ggplot2)
library(reshape2)
library(broom)
library(gridExtra)
library(car)  # For Levene's test
```

# Load and Prepare Data

```{r load-data}
# Read the SWISS.DAT file
swiss_data <- read.table("Rosner_FundamentalsBiostatistics_8e_Data_Sets/ASCII-tab/SWISS.DAT.txt",
                         header = TRUE,
                         sep = "\t",
                         quote = "'\"",
                         strip.white = TRUE)

# Display structure
str(swiss_data)

cat("\nDataset dimensions:", nrow(swiss_data), "subjects x", ncol(swiss_data), "variables\n")

# Create descriptive group labels
swiss_data <- swiss_data %>%
  dplyr::mutate(
    group_label = factor(group,
                        levels = c(1, 2, 3),
                        labels = c("High-NAPAP", "Low-NAPAP", "Control"))
  )

cat("\nSample sizes by group:\n")
table(swiss_data$group_label)

# Replace missing value code (9.99) with NA
swiss_data <- swiss_data %>%
  dplyr::mutate(across(starts_with("creat_"), ~ifelse(. == 9.99, NA, .)))

cat("\nMissing data summary by year:\n")
missing_summary <- swiss_data %>%
  dplyr::select(starts_with("creat_")) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Year", values_to = "N_Missing")

kable(missing_summary, caption = "Number of Missing Serum Creatinine Values by Year")
```

# Problem 12.47: ANOVA for Change from Baseline to 1978

## Calculate Change Scores

```{r problem-12-47-prep}
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
cat("PROBLEM 12.47: ANOVA FOR CHANGE IN SERUM CREATININE (1968 to 1978)\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n\n")

# Calculate change from baseline (1968) to 1978
swiss_change <- swiss_data %>%
  dplyr::mutate(
    baseline = creat_68,
    final = creat_78,
    change = creat_78 - creat_68
  ) %>%
  dplyr::select(Id, age, group, group_label, baseline, final, change) %>%
  dplyr::filter(!is.na(baseline) & !is.na(final))

cat("Sample sizes after removing cases with missing baseline or final values:\n")
table(swiss_change$group_label)
```

## Descriptive Statistics

```{r problem-12-47-descriptive}
# Descriptive statistics by group
change_stats <- swiss_change %>%
  dplyr::group_by(group_label) %>%
  dplyr::summarise(
    N = n(),
    Mean_Baseline = mean(baseline),
    SD_Baseline = sd(baseline),
    Mean_Final = mean(final),
    SD_Final = sd(final),
    Mean_Change = mean(change),
    SD_Change = sd(change),
    SE_Change = sd(change) / sqrt(n()),
    .groups = 'drop'
  )

kable(change_stats, digits = 3,
      caption = "Descriptive Statistics: Serum Creatinine Change from 1968 to 1978")

# Overall statistics
cat("\nOverall Change Statistics:\n")
cat("Mean change:", round(mean(swiss_change$change), 3), "\n")
cat("SD change:", round(sd(swiss_change$change), 3), "\n")
cat("Range:", round(range(swiss_change$change), 3), "\n")
```

## Visualizations

```{r problem-12-47-plots}
# Boxplot of change by group
ggplot(swiss_change, aes(x = group_label, y = change, fill = group_label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 2) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4,
               fill = "red", color = "darkred") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue", linewidth = 1) +
  labs(title = "Change in Serum Creatinine from 1968 to 1978 by Group",
       subtitle = "Red diamonds indicate group means; dashed line at zero",
       x = "NAPAP Exposure Group",
       y = "Change in Serum Creatinine (mg/dL)",
       fill = "Group") +
  theme_minimal() +
  theme(legend.position = "none")

# Before-After plot
change_long <- swiss_change %>%
  dplyr::select(Id, group_label, baseline, final) %>%
  pivot_longer(cols = c(baseline, final),
               names_to = "Time",
               values_to = "Creatinine") %>%
  dplyr::mutate(Time = factor(Time, levels = c("baseline", "final"),
                             labels = c("1968 (Baseline)", "1978 (Final)")))

ggplot(change_long, aes(x = Time, y = Creatinine, group = Id, color = group_label)) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.5, size = 1) +
  stat_summary(aes(group = group_label), fun = mean, geom = "line",
               linewidth = 2, alpha = 0.8) +
  stat_summary(aes(group = group_label), fun = mean, geom = "point",
               size = 4, shape = 18) +
  facet_wrap(~ group_label) +
  labs(title = "Individual and Mean Serum Creatinine: Baseline to Final",
       subtitle = "Thick lines and diamonds show group means",
       x = "Time Point",
       y = "Serum Creatinine (mg/dL)",
       color = "Group") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## One-Way ANOVA

```{r problem-12-47-anova}
# Check assumptions
cat("\n### Assumption Checks ###\n\n")

# Normality by group
cat("Shapiro-Wilk Tests for Normality of Change Scores:\n")
for (grp in levels(swiss_change$group_label)) {
  grp_data <- swiss_change %>% dplyr::filter(group_label == grp)
  shapiro_test <- shapiro.test(grp_data$change)
  cat(grp, ": W =", round(shapiro_test$statistic, 4),
      ", p =", format.pval(shapiro_test$p.value, digits = 3), "\n")
}

# Levene's test for homogeneity of variance
cat("\nLevene's Test for Homogeneity of Variance:\n")
levene_result <- leveneTest(change ~ group_label, data = swiss_change)
print(levene_result)

# Perform one-way ANOVA
cat("\n\n### One-Way ANOVA Results ###\n\n")
anova_change <- aov(change ~ group_label, data = swiss_change)
anova_summary <- summary(anova_change)
print(anova_summary)

# Extract key statistics
f_stat <- anova_summary[[1]]$`F value`[1]
p_val <- anova_summary[[1]]$`Pr(>F)`[1]
df_group <- anova_summary[[1]]$Df[1]
df_error <- anova_summary[[1]]$Df[2]

cat("\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
cat("ANOVA INTERPRETATION:\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
cat("F-statistic:", round(f_stat, 3), "\n")
cat("Degrees of freedom:", df_group, "and", df_error, "\n")
cat("p-value:", format.pval(p_val, digits = 4), "\n\n")

if (p_val < 0.05) {
  cat("CONCLUSION: There ARE statistically significant differences in mean\n")
  cat("serum creatinine change among the three groups (p < 0.05).\n")
} else {
  cat("CONCLUSION: There are NO statistically significant differences in mean\n")
  cat("serum creatinine change among the three groups (p >= 0.05).\n")
}

# Effect size
ss_group <- anova_summary[[1]]$`Sum Sq`[1]
ss_total <- sum(anova_summary[[1]]$`Sum Sq`)
eta_squared <- ss_group / ss_total
cat("\nEffect size (η²):", round(eta_squared, 4))
cat("\n", round(eta_squared * 100, 2), "% of variance explained by group\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")

# Tukey's HSD for pairwise comparisons
cat("\n\n### Post-hoc Pairwise Comparisons (Tukey's HSD) ###\n\n")
tukey_change <- TukeyHSD(anova_change)
print(tukey_change)

# Create summary table
tukey_df <- as.data.frame(tukey_change$group_label)
tukey_df$Comparison <- rownames(tukey_df)
tukey_summary <- tukey_df %>%
  dplyr::select(Comparison, diff, lwr, upr, `p adj`)

kable(tukey_summary, digits = 4,
      caption = "Tukey's HSD Pairwise Comparisons of Mean Change",
      col.names = c("Comparison", "Mean Difference", "95% CI Lower",
                    "95% CI Upper", "Adjusted p-value"))
```

# Problem 12.48: Individual Regression Lines

```{r problem-12-48}
cat("\n\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
cat("PROBLEM 12.48: INDIVIDUAL REGRESSION LINES (Creatinine vs. Time)\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n\n")

# Reshape data to long format
swiss_long <- swiss_data %>%
  dplyr::select(Id, age, group, group_label, starts_with("creat_")) %>%
  pivot_longer(cols = starts_with("creat_"),
               names_to = "visit_year",
               values_to = "creatinine") %>%
  dplyr::mutate(
    year = as.numeric(gsub("creat_", "19", visit_year)),
    time_from_baseline = year - 1968
  ) %>%
  dplyr::filter(!is.na(creatinine))

cat("Total observations in long format:", nrow(swiss_long), "\n")
cat("Observations per subject range from",
    min(table(swiss_long$Id)), "to", max(table(swiss_long$Id)), "\n\n")

# Fit individual regression lines for each subject
individual_models <- swiss_long %>%
  dplyr::group_by(Id, group, group_label) %>%
  dplyr::filter(n() >= 2) %>%  # Need at least 2 points for regression
  dplyr::do(model = lm(creatinine ~ time_from_baseline, data = .)) %>%
  dplyr::mutate(
    intercept = coef(model)[1],
    slope = coef(model)[2],
    n_obs = nobs(model)
  ) %>%
  dplyr::select(-model) %>%
  dplyr::ungroup()

cat("Number of subjects with regression lines:", nrow(individual_models), "\n")

# Summary statistics of slopes and intercepts by group
regression_summary <- individual_models %>%
  dplyr::group_by(group_label) %>%
  dplyr::summarise(
    N = n(),
    Mean_Intercept = mean(intercept),
    SD_Intercept = sd(intercept),
    Mean_Slope = mean(slope),
    SD_Slope = sd(slope),
    Median_Slope = median(slope),
    Min_Slope = min(slope),
    Max_Slope = max(slope),
    .groups = 'drop'
  )

kable(regression_summary, digits = 4,
      caption = "Summary of Individual Regression Parameters by Group")

cat("\n\nINTERPRETATION:\n")
cat("---------------\n")
cat("INTERCEPT: Represents the estimated serum creatinine level at baseline (1968)\n")
cat("           for each individual.\n\n")
cat("SLOPE: Represents the annual rate of change in serum creatinine (mg/dL per year)\n")
cat("       for each individual over the study period.\n")
cat("       - Positive slope: creatinine increasing over time (declining renal function)\n")
cat("       - Negative slope: creatinine decreasing over time (improving renal function)\n")
cat("       - Zero slope: no change in renal function over time\n\n")

# Display examples
cat("Examples of individual regression parameters:\n")
example_subjects <- individual_models %>%
  dplyr::group_by(group_label) %>%
  dplyr::slice_head(n = 3)

kable(example_subjects, digits = 4,
      caption = "Example Individual Regression Parameters (first 3 subjects per group)")
```

## Visualization of Individual Trajectories

```{r problem-12-48-plots, fig.height=10}
# Plot individual trajectories by group
ggplot(swiss_long %>% dplyr::filter(Id %in% individual_models$Id),
       aes(x = time_from_baseline, y = creatinine, group = Id)) +
  geom_line(alpha = 0.3, color = "gray50") +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(aes(group = group_label), method = "lm",
              se = TRUE, color = "red", linewidth = 2) +
  facet_wrap(~ group_label, ncol = 1) +
  labs(title = "Individual Serum Creatinine Trajectories by Group",
       subtitle = "Red line shows group-average trend with 95% CI",
       x = "Years from Baseline (1968)",
       y = "Serum Creatinine (mg/dL)") +
  theme_minimal()

# Distribution of slopes and intercepts
p1 <- ggplot(individual_models, aes(x = group_label, y = slope, fill = group_label)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Distribution of Individual Slopes",
       x = "Group",
       y = "Slope (mg/dL per year)") +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(individual_models, aes(x = group_label, y = intercept, fill = group_label)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Distribution of Individual Intercepts",
       x = "Group",
       y = "Intercept (baseline creatinine, mg/dL)") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p1, p2, ncol = 2)
```

# Problem 12.49: Compare Slopes Among Groups

```{r problem-12-49}
cat("\n\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
cat("PROBLEM 12.49: ANOVA COMPARING SLOPES AMONG GROUPS\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n\n")

# ANOVA for slopes
anova_slopes <- aov(slope ~ group_label, data = individual_models)
anova_slopes_summary <- summary(anova_slopes)
print(anova_slopes_summary)

# Extract statistics
f_slope <- anova_slopes_summary[[1]]$`F value`[1]
p_slope <- anova_slopes_summary[[1]]$`Pr(>F)`[1]

cat("\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
cat("INTERPRETATION - COMPARISON OF SLOPES:\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
cat("F-statistic:", round(f_slope, 3), "\n")
cat("p-value:", format.pval(p_slope, digits = 4), "\n\n")

if (p_slope < 0.05) {
  cat("CONCLUSION: The rate of change in serum creatinine (slopes) DIFFERS\n")
  cat("significantly among the three groups (p < 0.05).\n")
} else {
  cat("CONCLUSION: The rate of change in serum creatinine (slopes) does NOT differ\n")
  cat("significantly among the three groups (p >= 0.05).\n")
}
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")

# Tukey's HSD for slopes
cat("\n\n### Pairwise Comparisons of Slopes (Tukey's HSD) ###\n\n")
tukey_slopes <- TukeyHSD(anova_slopes)
print(tukey_slopes)

tukey_slopes_df <- as.data.frame(tukey_slopes$group_label)
tukey_slopes_df$Comparison <- rownames(tukey_slopes_df)

kable(tukey_slopes_df %>% dplyr::select(Comparison, diff, lwr, upr, `p adj`),
      digits = 4,
      caption = "Pairwise Comparisons of Mean Slopes",
      col.names = c("Comparison", "Difference in Slopes", "95% CI Lower",
                    "95% CI Upper", "Adjusted p-value"))
```

# Problem 12.50: Compare Intercepts Among Groups

```{r problem-12-50}
cat("\n\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
cat("PROBLEM 12.50: ANOVA COMPARING INTERCEPTS AMONG GROUPS\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n\n")

# ANOVA for intercepts
anova_intercepts <- aov(intercept ~ group_label, data = individual_models)
anova_intercepts_summary <- summary(anova_intercepts)
print(anova_intercepts_summary)

# Extract statistics
f_intercept <- anova_intercepts_summary[[1]]$`F value`[1]
p_intercept <- anova_intercepts_summary[[1]]$`Pr(>F)`[1]

cat("\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
cat("INTERPRETATION - COMPARISON OF INTERCEPTS:\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
cat("F-statistic:", round(f_intercept, 3), "\n")
cat("p-value:", format.pval(p_intercept, digits = 4), "\n\n")

if (p_intercept < 0.05) {
  cat("CONCLUSION: Baseline serum creatinine levels (intercepts) DIFFER\n")
  cat("significantly among the three groups (p < 0.05).\n")
} else {
  cat("CONCLUSION: Baseline serum creatinine levels (intercepts) do NOT differ\n")
  cat("significantly among the three groups (p >= 0.05).\n")
}
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")

# Tukey's HSD for intercepts
cat("\n\n### Pairwise Comparisons of Intercepts (Tukey's HSD) ###\n\n")
tukey_intercepts <- TukeyHSD(anova_intercepts)
print(tukey_intercepts)

tukey_intercepts_df <- as.data.frame(tukey_intercepts$group_label)
tukey_intercepts_df$Comparison <- rownames(tukey_intercepts_df)

kable(tukey_intercepts_df %>% dplyr::select(Comparison, diff, lwr, upr, `p adj`),
      digits = 4,
      caption = "Pairwise Comparisons of Mean Intercepts",
      col.names = c("Comparison", "Difference in Intercepts", "95% CI Lower",
                    "95% CI Upper", "Adjusted p-value"))
```

# Problem 12.51: Overall Conclusions

```{r problem-12-51}
cat("\n\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
cat("PROBLEM 12.51: OVERALL CONCLUSIONS\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n\n")

cat("SUMMARY OF FINDINGS:\n")
cat("===================\n\n")

cat("1. CHANGE FROM BASELINE TO 1978 (Problem 12.47):\n")
cat("   -----------------------------------------------\n")
cat("   Method: One-way ANOVA comparing mean change scores\n")
cat("   F-statistic:", round(f_stat, 3), "\n")
cat("   p-value:", format.pval(p_val, digits = 4), "\n")
cat("   Result:", ifelse(p_val < 0.05, "SIGNIFICANT", "NOT SIGNIFICANT"), "\n")
if (p_val < 0.05) {
  cat("\n   Mean changes by group:\n")
  for (i in 1:nrow(change_stats)) {
    cat("   ", change_stats$group_label[i], ":",
        sprintf("%.4f ± %.4f mg/dL",
                change_stats$Mean_Change[i],
                change_stats$SD_Change[i]), "\n")
  }
}
cat("\n")

cat("2. INDIVIDUAL REGRESSION ANALYSIS (Problem 12.48):\n")
cat("   ------------------------------------------------\n")
cat("   ", nrow(individual_models), "subjects had sufficient data for regression\n")
cat("   Mean slopes (rate of change per year):\n")
for (i in 1:nrow(regression_summary)) {
  cat("   ", regression_summary$group_label[i], ":",
      sprintf("%.4f ± %.4f mg/dL/year",
              regression_summary$Mean_Slope[i],
              regression_summary$SD_Slope[i]), "\n")
}
cat("\n")

cat("3. COMPARISON OF SLOPES (Problem 12.49):\n")
cat("   --------------------------------------\n")
cat("   Method: ANOVA comparing individual slopes\n")
cat("   F-statistic:", round(f_slope, 3), "\n")
cat("   p-value:", format.pval(p_slope, digits = 4), "\n")
cat("   Result:", ifelse(p_slope < 0.05, "SIGNIFICANT", "NOT SIGNIFICANT"), "\n")
cat("   Interpretation: The rate of change in serum creatinine",
    ifelse(p_slope < 0.05, "DIFFERS", "does NOT differ"),
    "among groups.\n\n")

cat("4. COMPARISON OF INTERCEPTS (Problem 12.50):\n")
cat("   ------------------------------------------\n")
cat("   Method: ANOVA comparing individual intercepts\n")
cat("   F-statistic:", round(f_intercept, 3), "\n")
cat("   p-value:", format.pval(p_intercept, digits = 4), "\n")
cat("   Result:", ifelse(p_intercept < 0.05, "SIGNIFICANT", "NOT SIGNIFICANT"), "\n")
cat("   Interpretation: Baseline creatinine levels",
    ifelse(p_intercept < 0.05, "DIFFER", "do NOT differ"),
    "among groups.\n\n")

cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
cat("OVERALL CLINICAL INTERPRETATION:\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n\n")

cat("This longitudinal study examined the effect of NAPAP exposure on renal function\n")
cat("as measured by serum creatinine levels over a 10-year period (1968-1978).\n\n")

cat("Key findings:\n")
cat("- Simple comparison of change scores (Problem 12.47) may be limited by only\n")
cat("  using first and last observations.\n\n")
cat("- Individual regression analysis (Problem 12.48) utilized all available data\n")
cat("  points for each subject, providing more robust estimates of change.\n\n")

if (p_slope < 0.05) {
  cat("- NAPAP exposure appears to affect the RATE of change in renal function,\n")
  cat("  with different groups showing different trajectories over time.\n\n")
} else {
  cat("- No significant difference in the rate of change (slopes) suggests that\n")
  cat("  NAPAP exposure does not substantially alter renal function trajectories.\n\n")
}

if (p_intercept < 0.05) {
  cat("- Baseline differences among groups suggest potential confounding or\n")
  cat("  pre-existing renal function differences.\n\n")
} else {
  cat("- Groups started with similar baseline renal function (similar intercepts).\n\n")
}

cat("\nLIMITATIONS:\n")
cat("- This analysis treats all subjects equally regardless of number of visits\n")
cat("- More advanced longitudinal methods (mixed models, GEE) would provide better\n")
cat("  estimates by accounting for within-subject correlation and varying visit patterns\n")
cat("- Missing data patterns should be examined for potential bias\n")

cat("\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
```

# Comparison of Methods

```{r method-comparison}
cat("\n\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n")
cat("COMPARISON OF ANALYTICAL APPROACHES\n")
cat("=" %>% rep(70) %>% paste(collapse = ""), "\n\n")

comparison_summary <- data.frame(
  Method = c("Change Score ANOVA (12.47)",
             "Slope Comparison (12.49)",
             "Intercept Comparison (12.50)"),

  Data_Used = c("Baseline and 1978 only",
                "All available time points",
                "All available time points"),

  F_Statistic = c(round(f_stat, 3),
                  round(f_slope, 3),
                  round(f_intercept, 3)),

  p_value = c(format.pval(p_val, digits = 4),
              format.pval(p_slope, digits = 4),
              format.pval(p_intercept, digits = 4)),

  Significant = c(ifelse(p_val < 0.05, "Yes", "No"),
                  ifelse(p_slope < 0.05, "Yes", "No"),
                  ifelse(p_intercept < 0.05, "Yes", "No")),

  Advantage = c("Simple, easy to interpret",
                "Uses all data, individual trajectories",
                "Assesses baseline differences")
)

kable(comparison_summary,
      caption = "Comparison of Different Analytical Approaches")
```

# Session Information

```{r session-info}
sessionInfo()
```
