---
title: "Problems 11.35-11.36: Hepatic Disease - Hormone Dose-Response Analysis"
author: "Linear Regression Analysis of Biliary and Pancreatic Secretions"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
    theme: cerulean
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  fig.align = 'center'
)
```

# Introduction

## Study Background

This document analyzes **dose-response relationships** for hepatic disease using the **HORMONE.DAT** dataset. The study examines how different hormone doses affect biliary and pancreatic function in hens.

### Research Questions

**Problem 11.35:** Use linear regression to assess dose-response relationships for:
- **Biliary secretion levels**
- **Pancreatic secretion levels**
- Separate analyses for each of the four active hormones

**Problem 11.36:** Use linear regression to assess dose-response relationships for:
- **Biliary pH levels**
- **Pancreatic pH levels**
- Separate analyses for each of the four active hormones

### Active Hormones Tested

The four active hormones tested are:
1. **Secretin**
2. **Cholecystokinin (CCK)**
3. **Gastrin**
4. **Glucagon**

# Data Loading and Preparation

```{r load-packages}
# Load required libraries
library(tidyverse)
library(knitr)
library(kableExtra)
library(car)         # For VIF and diagnostic plots
library(lmtest)      # For diagnostic tests
library(ggplot2)
library(gridExtra)   # For multiple plots
library(broom)       # For tidy model outputs

# Set theme for plots
theme_set(theme_minimal())
```



```{r load-data}
# Load the HORMONE.DAT dataset
data_path <- "Rosner_FundamentalsBiostatistics_8e_Data_Sets/ASCII-comma/HORMONE.DAT.txt"

# Read the data - handle quotes in column names
hormone_data <- tryCatch({
  read.csv(data_path, header = TRUE, strip.white = TRUE, quote = "'\"")
}, error = function(e) {
  cat("Error loading data from primary path. Trying alternative path...\n")
  read.csv("../Rosner_FundamentalsBiostatistics_8e_Data_Sets/ASCII-comma/HORMONE.DAT.txt",
           header = TRUE, strip.white = TRUE, quote = "'\"")
})

# Clean column names (remove any quotes)
colnames(hormone_data) <- gsub("'", "", colnames(hormone_data))
colnames(hormone_data) <- gsub("\"", "", colnames(hormone_data))

# Display structure
cat("Dataset Structure:\n")
str(hormone_data)

cat("\nColumn names in dataset:\n")
print(colnames(hormone_data))

cat("\nFirst few rows:\n")
head(hormone_data) %>%
  kable(caption = "First 6 Observations from HORMONE Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

# Variable Definitions

```{r variable-definitions}
# Create variable definitions table
var_definitions <- data.frame(
  Variable = c("ID", "Dose", "Hormone", "Bilsecpr", "Bilphpr", "Pansecpr", "Panphpr",
               "Bilsecpt", "Bilphpt", "Pansecpt", "Panphpt"),
  Description = c(
    "Hen identification number",
    "Dose level of hormone (μg/kg/hr)",
    "Type of hormone administered (1=Secretin, 2=CCK, 3=Gastrin, 4=Glucagon)",
    "Biliary secretion rate - pre-treatment (mL/hr)",
    "Biliary pH - pre-treatment",
    "Pancreatic secretion rate - pre-treatment (mL/hr)",
    "Pancreatic pH - pre-treatment",
    "Biliary secretion rate - post-treatment (mL/hr)",
    "Biliary pH - post-treatment",
    "Pancreatic secretion rate - post-treatment (mL/hr)",
    "Pancreatic pH - post-treatment"
  ),
  Type = c("ID", "Continuous (Predictor)", "Categorical",
           "Continuous", "Continuous", "Continuous", "Continuous",
           "Continuous (Outcome)", "Continuous (Outcome)",
           "Continuous (Outcome)", "Continuous (Outcome)")
)

kable(var_definitions,
      caption = "Variable Definitions for HORMONE Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

# Data Processing

```{r data-processing}
# Check for required columns
cat("Checking for required columns...\n")

# Display available columns
cat("\nAvailable columns:\n")
print(colnames(hormone_data))

# Process data
hormone_clean <- hormone_data %>%
  mutate(
    # Create hormone factor with meaningful labels
    Hormone_Label = factor(Hormone,
                          levels = c(1, 2, 3, 4),
                          labels = c("Secretin", "CCK", "Gastrin", "Glucagon")),

    # Rename columns for easier access (use post-treatment values)
    Biliary_Secretion = Bilsecpt,
    Biliary_pH = Bilphpt,
    Pancreatic_Secretion = Pansecpt,
    Pancreatic_pH = Panphpt
  )

# Remove control group (Dose = 0) for dose-response analysis
hormone_active <- hormone_clean %>%
  filter(Dose > 0)

cat("\nDataset summary:\n")
cat("  Total observations:", nrow(hormone_data), "\n")
cat("  Active hormone observations (Dose > 0):", nrow(hormone_active), "\n")
cat("  Control observations (Dose = 0):", sum(hormone_data$Dose == 0), "\n\n")

# Summary by hormone
cat("Observations by Hormone Type:\n")
hormone_summary <- hormone_active %>%
  group_by(Hormone_Label) %>%
  summarise(
    N = n(),
    Mean_Dose = round(mean(Dose, na.rm = TRUE), 2),
    SD_Dose = round(sd(Dose, na.rm = TRUE), 2),
    Min_Dose = round(min(Dose, na.rm = TRUE), 2),
    Max_Dose = round(max(Dose, na.rm = TRUE), 2)
  )

kable(hormone_summary,
      caption = "Summary Statistics by Hormone Type",
      col.names = c("Hormone", "N", "Mean Dose", "SD Dose", "Min Dose", "Max Dose")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)

cat("\nData processing completed successfully.\n")
```

# Exploratory Data Analysis

## Outcome Variables Summary

```{r outcome-summary}
# Summary statistics for outcome variables
outcome_summary <- hormone_active %>%
  summarise(
    N = n(),
    Mean_Biliary_Secretion = round(mean(Biliary_Secretion, na.rm = TRUE), 2),
    SD_Biliary_Secretion = round(sd(Biliary_Secretion, na.rm = TRUE), 2),
    Mean_Pancreatic_Secretion = round(mean(Pancreatic_Secretion, na.rm = TRUE), 2),
    SD_Pancreatic_Secretion = round(sd(Pancreatic_Secretion, na.rm = TRUE), 2),
    Mean_Biliary_pH = round(mean(Biliary_pH, na.rm = TRUE), 2),
    SD_Biliary_pH = round(sd(Biliary_pH, na.rm = TRUE), 2),
    Mean_Pancreatic_pH = round(mean(Pancreatic_pH, na.rm = TRUE), 2),
    SD_Pancreatic_pH = round(sd(Pancreatic_pH, na.rm = TRUE), 2)
  )

kable(t(outcome_summary),
      caption = "Overall Summary Statistics for Outcome Variables",
      col.names = "Value") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

## Summary by Hormone Type

```{r summary-by-hormone}
# Detailed summary by hormone
hormone_outcomes <- hormone_active %>%
  group_by(Hormone_Label) %>%
  summarise(
    N = n(),
    Mean_Biliary_Sec = round(mean(Biliary_Secretion, na.rm = TRUE), 2),
    SD_Biliary_Sec = round(sd(Biliary_Secretion, na.rm = TRUE), 2),
    Mean_Pancreatic_Sec = round(mean(Pancreatic_Secretion, na.rm = TRUE), 2),
    SD_Pancreatic_Sec = round(sd(Pancreatic_Secretion, na.rm = TRUE), 2),
    Mean_Biliary_pH = round(mean(Biliary_pH, na.rm = TRUE), 2),
    SD_Biliary_pH = round(sd(Biliary_pH, na.rm = TRUE), 2),
    Mean_Pancreatic_pH = round(mean(Pancreatic_pH, na.rm = TRUE), 2),
    SD_Pancreatic_pH = round(sd(Pancreatic_pH, na.rm = TRUE), 2)
  )

kable(hormone_outcomes,
      caption = "Outcome Variables by Hormone Type",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                font_size = 11)
```

## Distribution Visualizations

```{r distribution-plots}
# Biliary secretion by hormone
p1 <- ggplot(hormone_active, aes(x = Hormone_Label, y = Biliary_Secretion, fill = Hormone_Label)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Biliary Secretion by Hormone Type",
       x = "Hormone",
       y = "Biliary Secretion (mL/hr)") +
  theme_minimal() +
  theme(legend.position = "none")

# Pancreatic secretion by hormone
p2 <- ggplot(hormone_active, aes(x = Hormone_Label, y = Pancreatic_Secretion, fill = Hormone_Label)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Pancreatic Secretion by Hormone Type",
       x = "Hormone",
       y = "Pancreatic Secretion (mL/hr)") +
  theme_minimal() +
  theme(legend.position = "none")

# Biliary pH by hormone
p3 <- ggplot(hormone_active, aes(x = Hormone_Label, y = Biliary_pH, fill = Hormone_Label)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Biliary pH by Hormone Type",
       x = "Hormone",
       y = "Biliary pH") +
  theme_minimal() +
  theme(legend.position = "none")

# Pancreatic pH by hormone
p4 <- ggplot(hormone_active, aes(x = Hormone_Label, y = Pancreatic_pH, fill = Hormone_Label)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Pancreatic pH by Hormone Type",
       x = "Hormone",
       y = "Pancreatic pH") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

---

# Problem 11.35: Dose-Response for Secretion Levels

## Introduction

Assess dose-response relationships between hormone dose and secretion levels using linear regression:
- **Biliary secretion** = f(Dose)
- **Pancreatic secretion** = f(Dose)

Separate analyses for each of the four active hormones.

## Biliary Secretion Analysis

### Function to Perform Regression Analysis

```{r regression-function}
# Function to perform dose-response regression
perform_dose_response <- function(data, hormone_name, outcome_var, outcome_label) {

  cat("\n", strrep("=", 80), "\n", sep = "")
  cat("HORMONE:", hormone_name, "\n")
  cat("OUTCOME:", outcome_label, "\n")
  cat(strrep("=", 80), "\n\n")

  # Subset data for specific hormone
  subset_data <- data %>%
    filter(Hormone_Label == hormone_name)

  cat("Sample size:", nrow(subset_data), "\n\n")

  # Check if we have enough data
  if(nrow(subset_data) < 3) {
    cat("ERROR: Insufficient data for regression (n < 3)\n")
    return(NULL)
  }

  # Create formula
  formula_str <- paste(outcome_var, "~ Dose")

  # Fit linear regression model
  model <- lm(as.formula(formula_str), data = subset_data)

  # Print summary
  print(summary(model))

  # Tidy output
  tidy_results <- tidy(model, conf.int = TRUE)
  cat("\n=== Regression Coefficients ===\n")
  print(kable(tidy_results,
              caption = paste(hormone_name, "-", outcome_label),
              digits = 4) %>%
        kable_styling(bootstrap_options = c("striped", "hover"),
                      full_width = FALSE))

  # Model fit statistics
  glance_results <- glance(model)
  cat("\n=== Model Fit Statistics ===\n")
  cat(sprintf("  R² = %.4f\n", glance_results$r.squared))
  cat(sprintf("  Adjusted R² = %.4f\n", glance_results$adj.r.squared))
  cat(sprintf("  RMSE = %.4f\n", glance_results$sigma))
  cat(sprintf("  F-statistic = %.4f (df = %d, %d)\n",
              glance_results$statistic, 1, glance_results$df.residual))
  cat(sprintf("  P-value = %.4e\n", glance_results$p.value))

  # Interpretation
  slope <- tidy_results$estimate[2]
  slope_p <- tidy_results$p.value[2]

  cat("\n=== Interpretation ===\n")
  cat(sprintf("Slope: %.4f (95%% CI: %.4f to %.4f)\n",
              slope, tidy_results$conf.low[2], tidy_results$conf.high[2]))

  if(slope_p < 0.001) {
    cat(sprintf("*** HIGHLY SIGNIFICANT dose-response relationship (p < 0.001) ***\n"))
  } else if(slope_p < 0.01) {
    cat(sprintf("** SIGNIFICANT dose-response relationship (p < 0.01) **\n"))
  } else if(slope_p < 0.05) {
    cat(sprintf("* SIGNIFICANT dose-response relationship (p < 0.05) *\n"))
  } else {
    cat(sprintf("No significant dose-response relationship (p = %.4f)\n", slope_p))
  }

  cat(sprintf("\nFor each 1 μg/kg/hr increase in %s dose:\n", hormone_name))
  cat(sprintf("  %s %s by %.4f units on average\n",
              outcome_label,
              ifelse(slope > 0, "increases", "decreases"),
              abs(slope)))

  # Create scatterplot with regression line
  p <- ggplot(subset_data, aes(x = Dose, y = .data[[outcome_var]])) +
    geom_point(size = 3, alpha = 0.6, color = "#3498db") +
    geom_smooth(method = "lm", se = TRUE, color = "red", fill = "pink") +
    labs(title = paste("Dose-Response:", hormone_name, "-", outcome_label),
         subtitle = sprintf("y = %.4f + %.4f*x, R² = %.4f, p = %.4e",
                           tidy_results$estimate[1], slope,
                           glance_results$r.squared, glance_results$p.value),
         x = "Dose (μg/kg/hr)",
         y = outcome_label) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 14))

  print(p)

  # Diagnostic plots
  cat("\n=== Diagnostic Plots ===\n")
  par(mfrow = c(2, 2))
  plot(model, which = 1:4)
  par(mfrow = c(1, 1))

  # Return results
  return(list(
    hormone = hormone_name,
    outcome = outcome_label,
    model = model,
    tidy = tidy_results,
    glance = glance_results
  ))
}
```

### Secretin - Biliary Secretion

```{r secretin-biliary-sec}
result_sec_bil <- perform_dose_response(
  hormone_active,
  "Secretin",
  "Biliary_Secretion",
  "Biliary Secretion (mL/hr)"
)
```

### CCK - Biliary Secretion

```{r cck-biliary-sec}
result_cck_bil <- perform_dose_response(
  hormone_active,
  "CCK",
  "Biliary_Secretion",
  "Biliary Secretion (mL/hr)"
)
```

### Gastrin - Biliary Secretion

```{r gastrin-biliary-sec}
result_gas_bil <- perform_dose_response(
  hormone_active,
  "Gastrin",
  "Biliary_Secretion",
  "Biliary Secretion (mL/hr)"
)
```

### Glucagon - Biliary Secretion

```{r glucagon-biliary-sec}
result_glu_bil <- perform_dose_response(
  hormone_active,
  "Glucagon",
  "Biliary_Secretion",
  "Biliary Secretion (mL/hr)"
)
```

### Summary: Biliary Secretion Dose-Response

```{r summary-biliary-secretion}
cat("\n=== SUMMARY: BILIARY SECRETION DOSE-RESPONSE ===\n\n")

# Compile results
biliary_sec_summary <- data.frame(
  Hormone = c("Secretin", "CCK", "Gastrin", "Glucagon"),
  Slope = NA,
  Slope_SE = NA,
  CI_Lower = NA,
  CI_Upper = NA,
  P_value = NA,
  R_squared = NA,
  Significant = NA
)

results_list <- list(result_sec_bil, result_cck_bil, result_gas_bil, result_glu_bil)

for(i in 1:4) {
  if(!is.null(results_list[[i]])) {
    biliary_sec_summary$Slope[i] <- results_list[[i]]$tidy$estimate[2]
    biliary_sec_summary$Slope_SE[i] <- results_list[[i]]$tidy$std.error[2]
    biliary_sec_summary$CI_Lower[i] <- results_list[[i]]$tidy$conf.low[2]
    biliary_sec_summary$CI_Upper[i] <- results_list[[i]]$tidy$conf.high[2]
    biliary_sec_summary$P_value[i] <- results_list[[i]]$tidy$p.value[2]
    biliary_sec_summary$R_squared[i] <- results_list[[i]]$glance$r.squared
    biliary_sec_summary$Significant[i] <- ifelse(results_list[[i]]$tidy$p.value[2] < 0.05, "Yes", "No")
  }
}

kable(biliary_sec_summary,
      caption = "Summary of Dose-Response Relationships: Biliary Secretion",
      digits = 4,
      col.names = c("Hormone", "Slope", "SE", "95% CI Lower", "95% CI Upper",
                   "P-value", "R²", "Significant (α=0.05)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  row_spec(which(biliary_sec_summary$Significant == "Yes"),
           bold = TRUE, background = "#ffffcc")
```

## Pancreatic Secretion Analysis

### Secretin - Pancreatic Secretion

```{r secretin-pancreatic-sec}
result_sec_pan <- perform_dose_response(
  hormone_active,
  "Secretin",
  "Pancreatic_Secretion",
  "Pancreatic Secretion (mL/hr)"
)
```

### CCK - Pancreatic Secretion

```{r cck-pancreatic-sec}
result_cck_pan <- perform_dose_response(
  hormone_active,
  "CCK",
  "Pancreatic_Secretion",
  "Pancreatic Secretion (mL/hr)"
)
```

### Gastrin - Pancreatic Secretion

```{r gastrin-pancreatic-sec}
result_gas_pan <- perform_dose_response(
  hormone_active,
  "Gastrin",
  "Pancreatic_Secretion",
  "Pancreatic Secretion (mL/hr)"
)
```

### Glucagon - Pancreatic Secretion

```{r glucagon-pancreatic-sec}
result_glu_pan <- perform_dose_response(
  hormone_active,
  "Glucagon",
  "Pancreatic_Secretion",
  "Pancreatic Secretion (mL/hr)"
)
```

### Summary: Pancreatic Secretion Dose-Response

```{r summary-pancreatic-secretion}
cat("\n=== SUMMARY: PANCREATIC SECRETION DOSE-RESPONSE ===\n\n")

# Compile results
pancreatic_sec_summary <- data.frame(
  Hormone = c("Secretin", "CCK", "Gastrin", "Glucagon"),
  Slope = NA,
  Slope_SE = NA,
  CI_Lower = NA,
  CI_Upper = NA,
  P_value = NA,
  R_squared = NA,
  Significant = NA
)

results_list <- list(result_sec_pan, result_cck_pan, result_gas_pan, result_glu_pan)

for(i in 1:4) {
  if(!is.null(results_list[[i]])) {
    pancreatic_sec_summary$Slope[i] <- results_list[[i]]$tidy$estimate[2]
    pancreatic_sec_summary$Slope_SE[i] <- results_list[[i]]$tidy$std.error[2]
    pancreatic_sec_summary$CI_Lower[i] <- results_list[[i]]$tidy$conf.low[2]
    pancreatic_sec_summary$CI_Upper[i] <- results_list[[i]]$tidy$conf.high[2]
    pancreatic_sec_summary$P_value[i] <- results_list[[i]]$tidy$p.value[2]
    pancreatic_sec_summary$R_squared[i] <- results_list[[i]]$glance$r.squared
    pancreatic_sec_summary$Significant[i] <- ifelse(results_list[[i]]$tidy$p.value[2] < 0.05, "Yes", "No")
  }
}

kable(pancreatic_sec_summary,
      caption = "Summary of Dose-Response Relationships: Pancreatic Secretion",
      digits = 4,
      col.names = c("Hormone", "Slope", "SE", "95% CI Lower", "95% CI Upper",
                   "P-value", "R²", "Significant (α=0.05)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  row_spec(which(pancreatic_sec_summary$Significant == "Yes"),
           bold = TRUE, background = "#ffffcc")
```

---

# Problem 11.36: Dose-Response for pH Levels

## Introduction

Assess dose-response relationships between hormone dose and pH levels using linear regression:
- **Biliary pH** = f(Dose)
- **Pancreatic pH** = f(Dose)

Separate analyses for each of the four active hormones.

## Biliary pH Analysis

### Secretin - Biliary pH

```{r secretin-biliary-ph}
result_sec_bil_ph <- perform_dose_response(
  hormone_active,
  "Secretin",
  "Biliary_pH",
  "Biliary pH"
)
```

### CCK - Biliary pH

```{r cck-biliary-ph}
result_cck_bil_ph <- perform_dose_response(
  hormone_active,
  "CCK",
  "Biliary_pH",
  "Biliary pH"
)
```

### Gastrin - Biliary pH

```{r gastrin-biliary-ph}
result_gas_bil_ph <- perform_dose_response(
  hormone_active,
  "Gastrin",
  "Biliary_pH",
  "Biliary pH"
)
```

### Glucagon - Biliary pH

```{r glucagon-biliary-ph}
result_glu_bil_ph <- perform_dose_response(
  hormone_active,
  "Glucagon",
  "Biliary_pH",
  "Biliary pH"
)
```

### Summary: Biliary pH Dose-Response

```{r summary-biliary-ph}
cat("\n=== SUMMARY: BILIARY pH DOSE-RESPONSE ===\n\n")

# Compile results
biliary_ph_summary <- data.frame(
  Hormone = c("Secretin", "CCK", "Gastrin", "Glucagon"),
  Slope = NA,
  Slope_SE = NA,
  CI_Lower = NA,
  CI_Upper = NA,
  P_value = NA,
  R_squared = NA,
  Significant = NA
)

results_list <- list(result_sec_bil_ph, result_cck_bil_ph, result_gas_bil_ph, result_glu_bil_ph)

for(i in 1:4) {
  if(!is.null(results_list[[i]])) {
    biliary_ph_summary$Slope[i] <- results_list[[i]]$tidy$estimate[2]
    biliary_ph_summary$Slope_SE[i] <- results_list[[i]]$tidy$std.error[2]
    biliary_ph_summary$CI_Lower[i] <- results_list[[i]]$tidy$conf.low[2]
    biliary_ph_summary$CI_Upper[i] <- results_list[[i]]$tidy$conf.high[2]
    biliary_ph_summary$P_value[i] <- results_list[[i]]$tidy$p.value[2]
    biliary_ph_summary$R_squared[i] <- results_list[[i]]$glance$r.squared
    biliary_ph_summary$Significant[i] <- ifelse(results_list[[i]]$tidy$p.value[2] < 0.05, "Yes", "No")
  }
}

kable(biliary_ph_summary,
      caption = "Summary of Dose-Response Relationships: Biliary pH",
      digits = 4,
      col.names = c("Hormone", "Slope", "SE", "95% CI Lower", "95% CI Upper",
                   "P-value", "R²", "Significant (α=0.05)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  row_spec(which(biliary_ph_summary$Significant == "Yes"),
           bold = TRUE, background = "#ffffcc")
```

## Pancreatic pH Analysis

### Secretin - Pancreatic pH

```{r secretin-pancreatic-ph}
result_sec_pan_ph <- perform_dose_response(
  hormone_active,
  "Secretin",
  "Pancreatic_pH",
  "Pancreatic pH"
)
```

### CCK - Pancreatic pH

```{r cck-pancreatic-ph}
result_cck_pan_ph <- perform_dose_response(
  hormone_active,
  "CCK",
  "Pancreatic_pH",
  "Pancreatic pH"
)
```

### Gastrin - Pancreatic pH

```{r gastrin-pancreatic-ph}
result_gas_pan_ph <- perform_dose_response(
  hormone_active,
  "Gastrin",
  "Pancreatic_pH",
  "Pancreatic pH"
)
```

### Glucagon - Pancreatic pH

```{r glucagon-pancreatic-ph}
result_glu_pan_ph <- perform_dose_response(
  hormone_active,
  "Glucagon",
  "Pancreatic_pH",
  "Pancreatic pH"
)
```

### Summary: Pancreatic pH Dose-Response

```{r summary-pancreatic-ph}
cat("\n=== SUMMARY: PANCREATIC pH DOSE-RESPONSE ===\n\n")

# Compile results
pancreatic_ph_summary <- data.frame(
  Hormone = c("Secretin", "CCK", "Gastrin", "Glucagon"),
  Slope = NA,
  Slope_SE = NA,
  CI_Lower = NA,
  CI_Upper = NA,
  P_value = NA,
  R_squared = NA,
  Significant = NA
)

results_list <- list(result_sec_pan_ph, result_cck_pan_ph, result_gas_pan_ph, result_glu_pan_ph)

for(i in 1:4) {
  if(!is.null(results_list[[i]])) {
    pancreatic_ph_summary$Slope[i] <- results_list[[i]]$tidy$estimate[2]
    pancreatic_ph_summary$Slope_SE[i] <- results_list[[i]]$tidy$std.error[2]
    pancreatic_ph_summary$CI_Lower[i] <- results_list[[i]]$tidy$conf.low[2]
    pancreatic_ph_summary$CI_Upper[i] <- results_list[[i]]$tidy$conf.high[2]
    pancreatic_ph_summary$P_value[i] <- results_list[[i]]$tidy$p.value[2]
    pancreatic_ph_summary$R_squared[i] <- results_list[[i]]$glance$r.squared
    pancreatic_ph_summary$Significant[i] <- ifelse(results_list[[i]]$tidy$p.value[2] < 0.05, "Yes", "No")
  }
}

kable(pancreatic_ph_summary,
      caption = "Summary of Dose-Response Relationships: Pancreatic pH",
      digits = 4,
      col.names = c("Hormone", "Slope", "SE", "95% CI Lower", "95% CI Upper",
                   "P-value", "R²", "Significant (α=0.05)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  row_spec(which(pancreatic_ph_summary$Significant == "Yes"),
           bold = TRUE, background = "#ffffcc")
```

---

# Comprehensive Summary

## All Dose-Response Relationships

```{r comprehensive-summary}
cat("\n=== COMPREHENSIVE SUMMARY: ALL DOSE-RESPONSE RELATIONSHIPS ===\n\n")

# Create comprehensive summary table
comprehensive_summary <- data.frame(
  Outcome = rep(c("Biliary Secretion", "Pancreatic Secretion", "Biliary pH", "Pancreatic pH"), each = 4),
  Hormone = rep(c("Secretin", "CCK", "Gastrin", "Glucagon"), 4),
  Slope = c(biliary_sec_summary$Slope, pancreatic_sec_summary$Slope,
            biliary_ph_summary$Slope, pancreatic_ph_summary$Slope),
  P_value = c(biliary_sec_summary$P_value, pancreatic_sec_summary$P_value,
              biliary_ph_summary$P_value, pancreatic_ph_summary$P_value),
  R_squared = c(biliary_sec_summary$R_squared, pancreatic_sec_summary$R_squared,
                biliary_ph_summary$R_squared, pancreatic_ph_summary$R_squared),
  Significant = c(biliary_sec_summary$Significant, pancreatic_sec_summary$Significant,
                  biliary_ph_summary$Significant, pancreatic_ph_summary$Significant)
)

kable(comprehensive_summary,
      caption = "Comprehensive Summary of All Dose-Response Relationships",
      digits = 4,
      col.names = c("Outcome Variable", "Hormone", "Slope", "P-value", "R²", "Significant")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                font_size = 11) %>%
  row_spec(which(comprehensive_summary$Significant == "Yes"),
           bold = TRUE, background = "#ffffcc") %>%
  pack_rows("Problem 11.35: Secretion Levels", 1, 8) %>%
  pack_rows("Problem 11.36: pH Levels", 9, 16)
```

## Visual Summary: Slope Comparisons

```{r slope-comparison-plot}
# Create visualization of slopes
ggplot(comprehensive_summary, aes(x = Hormone, y = Slope, fill = Significant)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~Outcome, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("No" = "#95a5a6", "Yes" = "#2ecc71")) +
  labs(title = "Dose-Response Slopes by Hormone and Outcome",
       subtitle = "Green = Statistically Significant (p < 0.05)",
       x = "Hormone",
       y = "Slope (Change per 1 μg/kg/hr dose increase)",
       fill = "Significant") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold", size = 14))
```

## Significant Relationships Summary

```{r significant-summary}
cat("\n=== STATISTICALLY SIGNIFICANT DOSE-RESPONSE RELATIONSHIPS (p < 0.05) ===\n\n")

# Filter significant relationships
significant_relationships <- comprehensive_summary %>%
  filter(Significant == "Yes") %>%
  arrange(P_value)

if(nrow(significant_relationships) > 0) {
  kable(significant_relationships,
        caption = "Statistically Significant Dose-Response Relationships",
        digits = 4,
        row.names = FALSE,
        col.names = c("Outcome Variable", "Hormone", "Slope", "P-value", "R²", "Significant")) %>%
    kable_styling(bootstrap_options = c("striped", "hover"),
                  full_width = FALSE)

  cat("\n\nTotal significant relationships:", nrow(significant_relationships), "out of 16\n")
} else {
  cat("No statistically significant dose-response relationships detected at α = 0.05\n")
}
```

# Conclusions

## Problem 11.35: Secretion Levels

### Key Findings

**Biliary Secretion:**
```{r conclusions-biliary-sec, echo=FALSE, results='asis'}
for(i in 1:4) {
  hormone <- biliary_sec_summary$Hormone[i]
  sig <- biliary_sec_summary$Significant[i]
  slope <- biliary_sec_summary$Slope[i]
  pval <- biliary_sec_summary$P_value[i]

  if(!is.na(sig)) {
    if(sig == "Yes") {
      cat(sprintf("- **%s**: SIGNIFICANT dose-response (slope = %.4f, p = %.4f)\n",
                  hormone, slope, pval))
    } else {
      cat(sprintf("- **%s**: No significant dose-response (slope = %.4f, p = %.4f)\n",
                  hormone, slope, pval))
    }
  }
}
```

**Pancreatic Secretion:**
```{r conclusions-pancreatic-sec, echo=FALSE, results='asis'}
for(i in 1:4) {
  hormone <- pancreatic_sec_summary$Hormone[i]
  sig <- pancreatic_sec_summary$Significant[i]
  slope <- pancreatic_sec_summary$Slope[i]
  pval <- pancreatic_sec_summary$P_value[i]

  if(!is.na(sig)) {
    if(sig == "Yes") {
      cat(sprintf("- **%s**: SIGNIFICANT dose-response (slope = %.4f, p = %.4f)\n",
                  hormone, slope, pval))
    } else {
      cat(sprintf("- **%s**: No significant dose-response (slope = %.4f, p = %.4f)\n",
                  hormone, slope, pval))
    }
  }
}
```

## Problem 11.36: pH Levels

### Key Findings

**Biliary pH:**
```{r conclusions-biliary-ph, echo=FALSE, results='asis'}
for(i in 1:4) {
  hormone <- biliary_ph_summary$Hormone[i]
  sig <- biliary_ph_summary$Significant[i]
  slope <- biliary_ph_summary$Slope[i]
  pval <- biliary_ph_summary$P_value[i]

  if(!is.na(sig)) {
    if(sig == "Yes") {
      cat(sprintf("- **%s**: SIGNIFICANT dose-response (slope = %.4f, p = %.4f)\n",
                  hormone, slope, pval))
    } else {
      cat(sprintf("- **%s**: No significant dose-response (slope = %.4f, p = %.4f)\n",
                  hormone, slope, pval))
    }
  }
}
```

**Pancreatic pH:**
```{r conclusions-pancreatic-ph, echo=FALSE, results='asis'}
for(i in 1:4) {
  hormone <- pancreatic_ph_summary$Hormone[i]
  sig <- pancreatic_ph_summary$Significant[i]
  slope <- pancreatic_ph_summary$Slope[i]
  pval <- pancreatic_ph_summary$P_value[i]

  if(!is.na(sig)) {
    if(sig == "Yes") {
      cat(sprintf("- **%s**: SIGNIFICANT dose-response (slope = %.4f, p = %.4f)\n",
                  hormone, slope, pval))
    } else {
      cat(sprintf("- **%s**: No significant dose-response (slope = %.4f, p = %.4f)\n",
                  hormone, slope, pval))
    }
  }
}
```

## Biological Interpretation

The dose-response analysis reveals important patterns in how different hormones affect biliary and pancreatic function:

1. **Secretion Levels**: Hormones showing significant positive slopes indicate dose-dependent stimulation of secretion
2. **pH Levels**: Significant slopes indicate dose-dependent changes in acidity/alkalinity
3. **Hormone-Specific Effects**: Different hormones may have selective effects on different outcomes

## Statistical Methods Used

- **Simple Linear Regression** (Dose-Response models)
- **Coefficient Estimation** with 95% confidence intervals
- **Hypothesis Testing** for slope significance
- **Model Fit Assessment** (R², Adjusted R², RMSE)
- **Regression Diagnostics** (residual plots, assumption checking)

## Limitations

- Linear models assume constant dose-response relationship
- May not capture non-linear effects at extreme doses
- Sample size varies by hormone type
- Results apply to this specific experimental design (hen model)

# Session Information

```{r session-info}
sessionInfo()
```

---
**End of Analysis**
